Why do we Eval?\\[2mm]\Large A large-scale study of Eval usage in R

### 2020-12-19 paper/main.tex
usage rewrite

To provide a picture of \eval usage in R, this section focuses on
library code in the \Corpus packages of our CRAN corpus. For these,
all runnable code in CRAN was executed, runs that did not exercise an
\eval from the corpus were discarded; \Nbruns runs were retained and
generated the data presented here. In total, \Allcalls invocations of
\eval and its variants were recorded. Out of the
\Staticatleastonecallsite packages that use \eval, \Triggeredpkgs
packages were observed calling it --- the difference can be chalked
down to incomplete code coverage and occasional infrastructure
failures.

Figure~\ref{freq} summarizes the frequency of calls to \eval over all
runs. The left column is the number of observed calls, the right
column is the number of packages falling in that range. There are
\Fewcalls packages with low \eval frequency, i.e. fewer than 100
calls; and a similar number of packages, \Manycalls to be precise,
that use \eval more than 1,000 times. One package, \Maxcallspack,
calls it \Maxcalls times and thus accounts for over half of the
observed invocations of \eval.

@@

\begin{wrapfigure}{r}{5.8cm}
\vspace{-4mm}

  \begin{tabular}{|r@{\,}r@{\,}l@{\,}r|r@{\,}r@{\,}l@{}r|} \hline
\multicolumn{3}{|c}{\small\#calls} &\small\#sites &
\multicolumn{3}{c}{\small\#calls} &\small\#sites \\\hline
\tt 0 &--& \tt 50    & \Runbina & \tt 501 &--& \tt 1000   & \Runbine\\
\tt 51 &--& \tt 100  & \Runbinb & \tt 1001 &--& \tt 1500  & \Runbinf\\
\tt 101 &--& \tt 250 & \Runbinc & \tt 1501 &--& \tt 2000  & \Runbing\\
\tt 251 &--& \tt 500 & \Runbind & \tt 2001 &--& \tt 3000 & \Runbinh\\\hline
\end{tabular}
\caption{Normalized invocations} \label{tab:callsitesnormalized}

\vspace{-2mm}
\includegraphics[width=7.5cm, trim=7.5cm 0 6cm 0, clip]{calls_per_run_per_call_site}
\vspace{-7mm}
    \caption{Normalized invocations
      (small)} \label{fig:callsitesnormalized}
\vspace{-8mm}
\end{wrapfigure}

Figure~\ref{tab:variantseval} illustrates the use of variants of
\eval. For each of the four variants, the first rows shows the number
of call sites occurring in the corpus (\emph{static)}), the second row
is the number of call sites that were encountered during tracing
(\emph{exercised}), the last row is the number of calls
(\emph{invocations}). The vast majority of call sites, and calls, are
to \eval itself; \c{evalq} and \c{local} are barely used at all. The
difference between static and exercised reflects spotty code coverage.

Figure~\ref{tab:callsitesnormalized} shows normalized invocation
counts. The column on the left is the average number of invocations of
\eval from a particular call site for any given run. The number on the
right is number of call sites that fall in that range. For instance,
there are \Runbinh call sites that are invoked on average over 2,001
times in any run they are exercised.
Larger numbers of calls suggest that \eval may be called within a loop
or from a recursive context -- this seems to be the exception as the
majority of call sites, \Runbina, are exercised less than 51 times.
Figure~\ref{fig:callsitesnormalized} zooms in on low frequency call
sites. The x-axis shows the normalized invocations and the y-axis is
the number of call sites for that value. The figure shows that most
call sites are invoked only once, about half as many are invoked
twice. After that the frequency of invocation quickly drops.

Analysis of the arguments passed to \eval reveals that \Symbolpercent
are of kind \c{symbol} (these are single variables such as \c{x}),
\Languagepercent are \c{language} (an internal R type roughly
corresponding to a single expression, such as \c{x+1}), and
\Expressionpercent are \c{expression} (one or more statements).  These
account for \Codepercent of the invocations, the arguments to the
remaining invocations are R values which are returned unchanged by
\eval. Most of the invocations involving a symbol come from ggplot2, \Ggplotsymbolpercent, and from one call site, in \c{ggproto}, evaluating symbol \c{`_inherit`}.

To estimate how much executable code is injected into the program
through \eval, we measure the size of arguments in number of
AST nodes. For instance, a variable name (a symbol) counts for 1 node. Our de-parser does not emit code for values as some values
can be very large and would overwhelm the code. On average, arguments
to \eval are \Avgszeval nodes large, the median size is
\Medianszeval. The largest input observed is \Maxszeval.
Figure~\ref{fig:sizedistribution} shows the distribution of sizes. The
x-axis is the size of arguments in number of nodes, and the y-axis is the
count of arguments with that size.

\begin{figure}[h!]
 \includegraphics[width=\textwidth]{size_loaded_distribution}
\caption{Loaded code} \label{fig:sizedistribution}
\end{figure}

To measure the work performed in \evals, we instrument the R
interpreter to count the number of operations performed during
evaluation of the argument. For the purposes of this experiment, each
interpreter step is an \emph{event}. For our corpus, \EvalEventAllPerc
of events (\EventsRnd) are performed within \eval. \Smalleventspct of
\evals perform 50 or fewer events. Figure~\ref{fig:eventsPerPackage}
shows two violin plots. The plot on the left is for evals with fewer
or equal than 50 events, the plot on the right is for more
work-intensive \evals. The biggest \eval performs \MaxeventsRnd
events.

\begin{figure}[h!]
	\centering
	\includegraphics[width=.55\textwidth]{events_per_pack_small}
	\hspace{-10mm}
	\includegraphics[width=.55\textwidth]{events_per_pack_large}
	\caption{Events per packages} \label{fig:eventsPerPackage}
\end{figure}

\subsection{Things todo}

\begin{itemize}
\item check that on figure 7, the number of elements at 1 is correct
\item explain the peak at 4, for figure 7
\item Figure 8 explain why so few 1's
\item Add a ``Discussion'' paragraph at the end that sums up the
  observations
\item Add paragraphs for Kaggle and CORE -- only discuss the
  differences from CRAN if any.
\item Question: should we remove ggplot? Would change things
  significantly?
\end{itemize}

>>>>

This section focuses on our corpus of \Corpus CRAN packages that use
\eval; these represent {\bf XX\%} of the entire repository. To provide
a picture of \eval usage, all runnable code in CRAN was executed. Runs
that did not exercise an \eval in the corpus were discarded. We
retained \Nbruns to generate the data presented here. In total,
\Allcalls invocations of \eval and its variants were recorded. Out of
the \Staticatleastonecallsite packages that use \eval, \Triggeredpkgs
packages were observed calling it in our runs --- the difference can
be chalked down to incomplete code coverage and occasional
infrastructure failures.

Figure~\ref{freq} summarizes the frequency of calls to \eval over all
runs. The left column is the number of observed calls, the right
column is the number of packages falling in that range. There are
\Fewcalls packages with low \eval frequency, i.e. fewer than 100
calls; and a similar number of packages, \Manycalls to be precise,
that use \eval more than 1,000 times. One package, \Maxcallspack,
calls it \Maxcalls times and thus accounts for over half of the
observed data.

@@

\begin{wrapfigure}{r}{5.8cm}
\vspace{-4mm}
\centering
  \begin{tabular}{|r@{\,}r@{\,}l@{\,}r|r@{\,}r@{\,}l@{}r|} \hline
\multicolumn{3}{|c}{\small\#calls} &\small\#sites &
\multicolumn{3}{c}{\small\#calls} &\small\#sites \\\hline
\tt 0 &--& \tt 50    & \Runbina & \tt 501 &--& \tt 1000   & \Runbine\\
\tt 51 &--& \tt 100  & \Runbinb & \tt 1001 &--& \tt 1500  & \Runbinf\\
\tt 101 &--& \tt 250 & \Runbinc & \tt 1501 &--& \tt 2000  & \Runbing\\
\tt 251 &--& \tt 500 & \Runbind & \tt 2001 &--& \tt 3000 & \Runbinh\\\hline
\end{tabular}

\medskip  (a) All

  \includegraphics[width=5.8cm, trim=7.5cm 0 6cm 0, clip]{calls_per_run_per_call_site}

  (b) Small

\vspace{-2mm}
    \caption{Normalized invocations} \label{cn}
\vspace{-2mm}
\end{wrapfigure}

Figure~\ref{tab:variantseval} illustrates the use of variants of
\eval. For each of the four variants, the first rows shows the number
of call sites occurring in the corpus (\emph{static}), the second row
is the number of call sites that were encountered during tracing
(\emph{exercised}), the last row is the number of calls
(\emph{invocations}). The vast majority of call sites, and calls, are
to \eval itself, \c{eval.parent} is rare, as for \c{evalq} and
\c{local} are barely used at all. The difference between static and
exercised reflects spotty code coverage and is a limitation
of dynamic analysis.

Figure~\ref{cn}(a) shows normalized invocation counts. The column on
the left is the average number of calls from a particular site for any
given run (\emph{\#calls}). On the right is the count of sites that
fall in that range (\emph{\#sites}). For instance, \Runbinh sites are
invoked on average over 2,001 times per run. Larger numbers suggest
uses from within a loop or a recursive context -- this seems to be the
exception as most \evals, \Runbina, are exercised 50 times or less.
Figure~\ref{cn}(b) zooms in on low frequency call sites. The x-axis
shows normalized invocations and the y-axis is the number sites for
that value. Most low-frequency \evals are invoked only once, about
half as many are invoked twice; after that the frequency quickly
drops.

Arguments to \eval can be either code or values, in the latter case
\eval has no effect. Code accounts for \Codepercent of arguments in
our corpus. More specifically, \Symbolpercent of arguments are
\c{symbol}s (i.e. these are single variables such as \c{x}),
\Languagepercent are \c{language} object (an internal R type roughly
corresponding to a single expression, such as \c{x+1}), and
\Expressionpercent are \c{expression} objects (multiple statements).
Further inspection reveals that most symbols, \Ggplotsymbolpercent to
be exact, come from a single call in the \c{ggplot2} package and have
have the value \c{_inherit}. Even if we remove this outlier, the
majority of \evals are simple expressions.

To estimate how much executable code is dynamically injected into the
program through \eval, we measure the size of arguments in number of
abstract syntax tree nodes as returned by R's parser; for example,
\c{x+1} counts as 3 nodes.\footnote{We experimented with measuring
code size by de-parsing the expressions to strings, but these
measurements were dominated by de-parsed data objects, these could
range in the MBs.} The median argument size is \Medianszeval (recall
the large number of symbols) and the average is \Avgszeval nodes. The
largest \eval input observed is \Maxszeval\,-- a significant chunk of
code. Figure~\ref{fig:sizedistribution} shows the distribution of
sizes for arguments of fewer or equal to 25 nodes. The x-axis is the
size of arguments in number of nodes, and the y-axis is the count of
arguments with that size. The size drops rapidly, with few
observations larger than 15 nodes. The long tail is omitted from the
graph for legibility.

\begin{figure}[h!]
 \includegraphics[width=1.061\textwidth]{size_loaded_distribution}
\caption{Loaded code} \label{fig:sizedistribution}
\end{figure}

To measure the work performed in \evals, we count the instructions
executed by interpreter during evaluation of the argument. In our
corpus, \EvalEventAllPerc of all instructions (\EventsRnd) are
performed within \eval. Most invocations perform relatively little
work with \Smalleventspct of \evals executing 50 or fewer
instructions. The violin plot of Figure~\ref{ev}(a) corresponds to
\evals executing $\leq$ 50 instructions, it is dominated by trivial
symbol look ups. Figure~\ref{ev}(b) has the work-intensive \evals
which go all the way to \MaxeventsRnd instructions.

\begin{figure}[h!]
\begin{tabular}{@{}c@{}c@{}}
\begin{minipage}{7.5cm}
 \includegraphics[width=\textwidth]{events_per_pack_small}
\end{minipage}&\begin{minipage}{7.5cm}
  \includegraphics[width=\textwidth]{events_per_pack_large}
\end{minipage}\\[-3mm]
\small (a) Small & \small (b) Large
\end{tabular}
 \caption{Events per packages} \label{ev}
\end{figure}

\subsection{Things to do}

\begin{itemize}
\item Figure 5(b) -- make `count` and `average` smaller.
\item Figure 6: (1) Replace `ast-nodes` with `nodes`
\item Figure 6: (2) Remove gray lines
\item Figure 6: increase font size
\item Figure 6: Try to make the figure less wide. Thinner bars/aspect
  ratio change.
\item Figure 7: check that element counts are correct
\item Figure 7: replace events by instructions. Check what we are
  counting? Are these bytecodes or AST nodes.
\item Check that the percentage of operations performed within
  an \eval (5\%) is correct. That feels surprising to me.
\item I am having second thoughts about keeping the \c{ggplot2} data
  -- it is a big outlier that skews all the graphs...
\item Add a ``Discussion'' paragraph at the end that sums up the
  observations
\item Add paragraphs for Kaggle and CORE -- only discuss the
  differences from CRAN
\end{itemize}
